### 2013-10-24 src2/thesis.tex
update

@@

The approach taken to embedd the scalar language is very similar to what is described 
in~\citet{COMPILEEDSL}; but we use a GADT to obtained typed expressions rather than phantom types.

@@

\noindent Here we explore another way to provide the capabilities of ArBB to the 
Haskell programmer. Rather than trying to use ArBB as a backend to Accelerate 
we provide a more direct mapping of ArBB's functionality to Haskell idioms. 
Simply put, this means providing a more Haskell-programmer-friendly interface compared
to the raw ArBB bindings (which are a collection of functions in the IO monad).

>>>>

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
%
% GPU Programming 
%
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\section{GPUs and CUDA}

In 2006 NVIDIA released their first Compute Unified Device Architechture (CUDA) 
GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs 
for general purpose workloads. A dialect of C, called CUDA C, was released specifically 
to simplify implementation of general purpose algorithms on GPUs.

A CUDA GPU is built around a single kind of processor (opposed to the different 
kinds of processors found in earlier GPUs). The processors in the GPU (called MPs, 
MultiProcessors), all contain a number of function units called SPs (Streaming Processors). 
Each MP also contains local memory, called shared memory since it can be accessed 
by all of the SPs in that MP. The number of MPs vary over the available GPUs; cheaper 
GPUs have as few as one MP, and as you go up in price the number of MPs increase.

Each MP of the GPU can manage a large number of threads; on today's GPUs up to 
2048 threads can run on a single MP. The GPU schedules threads in groups of 32, called 
a {\em Warp}, that are executed in lock-step (SIMD style execution). Threads are 
also divided into {\em Blocks}; the threads within a block can communicate using the 
local shared memory. The maximum block size is currently 1024 threads. Since 
the number of threads in a block can be higher than the warp size, a barrier 
synchronisation mechanism exists to ensure that threads within a block 
can communicate safely via the shared memory.

Papers B, C, E and F (sections~\ref{sec:paperB},~\ref{sec:paperC},~\ref{sec:paperE},
~\ref{sec:paperF}) contain more information about GPUs, and the assumptions we make 
about them ,in relation to Obsdian.

\subsection{CUDA programming}

The CUDA programming model reflects the hierarchical architecture of GPUs. The programmer 
writes what is called {\em kernels}, in CUDA jargon a kernel refers to a sequential C program 
that is parameterised on a threads identity. The kernel program is executed in parallel across 
all the threads of a block and many such blocks of threads can execute in parallel on the 
available MPs. The programmer needs to decompose the algorithm and data in such a way what 
communication will be local to the blocks.

Below is a very simple CUDA example to give some indication of what a kernel and 
the code to launch that kernel looks like.

\begin{small}
\begin{verbatim} 
__global__ void kernel(int32_t *i1, int32_t *i2, int32_t *r) {

  r[threadIdx.x] = i1[threadIdx.x] + i2[threadIdx.x];

} 
\end{verbatim}
\end{small}

This kernel takes two input arrays, {\tt i1} and {\tt i2}, and computes 
elementwise sums into the result array, {\tt r}. {\tt threadIdx} identifies 
a thread within a block. There is also a {\tt blockIdx} variable that identifies 
the block a thread belongs to but that is not used in this case.

In CUDA there is a notion of {\em host} and {\em device}, the device is the GPU and 
its memory (Global memory, usually a few Gigabytes). The host refers to the system housing 
the GPU. The host runs a control program that launches kernels on the GPU. First memory 
needs to be allocated in the device to hold the inputs and outputs. Then data is copied from 
the host systems memory into device memory.

\begin{small}
\begin{verbatim} 
int main(void) {

  int32_t h_i1[32], h_i2[32], h_r[32];

  int32_t *d_i1, *d_i2, *d_r;

  for (int i = 0; i < 32; ++i) { 
    h_i1[i] = i;
    h_i2[i] = 32 - i;
  }

  // Allocate device memory
  cudaMalloc((void**)&d_i1,32*sizeof(int32_t));
  cudaMalloc((void**)&d_i2,32*sizeof(int32_t));
  cudaMalloc((void**)&d_r,32*sizeof(int32_t));

  // Copy inputs to device
  cudaMemcpy(d_i1,
             h_i1,
             32*sizeof(int32_t),
             cudaMemcpyHostToDevice);
  cudaMemcpy(d_i2,
             h_i2,
             32*sizeof(int32_t),
             cudaMemcpyHostToDevice);

  // Launch the kernel
  kernel<<<1,32,0>>>(d_i1,d_i2,d_r);

  // Copy result to host
  cudaMemcpy(h_r,
             d_r,
             32*sizeof(int32_t),
             cudaMemcpyDeviceToHost);

  for (int i = 0; i < 32; ++i) { 
    printf("%d ",h_r[i]);
  }

  return 0;
}
\end{verbatim} 
\end{small}

The syntax for launching a kernel is:

\begin{small}
\begin{verbatim}
name<<<#Blocks, #Threads/Block, #BytesShared>>>(arg1,..,argN);
\end{verbatim}
\end{small}

In this case one block of 32 threads using no shared memory, is launched.

@@

The approach taken to embedd the scalar language is very similar to what is described 
in~\citet{COMPILEEDSL}; but we use a Generalised Algebraic Datatype (GADT) to obtained 
typed expressions rather than phantom types.

@@

\noindent Here we explore another way to provide the capabilities of ArBB to the 
Haskell programmer. Rather than trying to use ArBB as a backend to Accelerate 
we provide a more direct mapping of ArBB's functionality to Haskell idioms. 
We call this embedded language EmbArBB and it provides a more Haskell-programmer-friendly 
interface compared to the raw ArBB bindings (which are a collection of functions 
in the IO monad).
